Chasing \rainbows{} and ocean glint:\\ Inner working angle constraints for the Habitable Worlds Observatory

### 2023-05-02 src/tex/ms.tex
[showyourwork] overleaf sync

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is the specular (Fresnel) reflection of light off smooth liquid surfaces. While glint is a familiar phenomenon on Earth, there is only
one other body in the Solar System where it has been observed: 
the Cassini orbiter captured the glint of sunlight reflecting off the
liquid hydrocarbon lakes near Titan's north pole as it was observing
the moon from the nightside \citep{2010GeoRL..37.7104S}.
%
In the total flux, the glint signal is indeed expected to be most prominent at large planetary phase angles.
%
The maximum polarization of the glint occurs at the Brewster angle, which is determined by the refractive indices at the interface; between air and liquid water this occurs at an angle of reflection of \qty{53}{\degree} \citep[i.e., at a planetary phase angle of \qty{127}{\degree}; see, e.g.,][]{2008Icar..195..927W}.
%
This angle therefore varies slightly with the wavelength and the composition of the ocean and the atmosphere.
%

In general, the glint on a planet will be strongest at red wavelengths where the Rayleigh scattering in the planet's atmosphere and attenuation through the atmosphere are smallest \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also small \citep{2012ApJ...752L...3C}.
%
The stronger glint signature at longer wavelengths has indeed been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
%
The width across phase angles of the glint feature in total and polarized flux generally
broadens with increasing wave height \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}, which depends strongly on the wind speed over the surface, as described by \citep{CoxMunk1954}.
%

The glint provides evidence of liquid surface bodies, which, taken in the context of the temperature and types of condensates, may be interpreted as evidence for the presence of liquid water.
%
At large phase angles, the forward scattering of light by cloud particles could be confused with glint \citep{Robinson_2010}, which further motivated probing the rainbow angle for context.
%
This confusion can be solved by measuring the intersection of phase curves at various wavelengths in polarized light (the polarized colour changes from blue, through white, to red), which only occurs when the glint contributes to the signal and at a planetary scattering angle depending on the cloud coverage fraction \citep{treesandstam2019}, as shown in Fig.~\ref{fig:bottplot}.
%The cloud coverage fraction on ocean worlds could be constrained with observations spanning the ultraviolet to the near-infrared by finding the 
%angle at which the planet's polarization phase curves at various wavelengths intersect \citep{treesandstam2019}.
%
While the glint feature can only be probed when a planet is at crescent, i.e.\ large phase angles, \citet{trees2022} showed that reflection by a liquid surface body can also be detected with spectropolarimetry, as the glint changes the difference in polarization between continuum wavelengths and absorption bands. This could be applied when the planet is near quadrature, and thus more easily probed when using coronagraphs.
%how use to map, 
%seen in Earthshine
%strong signal for exoplanets
% disentangled from cloud coupled with rainbow
%disentangled from lattitude albedo coupled in polarization

@@

Rainbows, cloudbows, glories and other optical phenomena from light that is 
scattered by atmospheric condensates can dramatically influence the shape of a planet's total flux and polarized phase curves and provide a wealth of information 
about exoplanets \citep{karalidi2012rainbow, stam2008,Bailey2007,2014A&A...566L...1G}.%
\footnote{
    Rainbows occur when light is scattered by rain droplets that are large compared to the wavelength of the light. Cloudbows refer to features occurring at the same phase angles from smaller cloud particles with the spread in colors invisible to the human eye making them appear white \citet[see][for further discussion]{Bailey2007}.  
}
%
In our own Solar System, the polarized back scatter from Titan provided evidence that the moon's reflection was due to an atmospheric haze rather than a surface \citep{zellner1973polarization}, the polarized phase curve of Venus at three wavelengths showed not only that the clouds were $\sim\qty{75}{\percent}$ H\textsubscript{2}SO\textsubscript{4}--water mixture but also constrained the droplet size distribution \citep{hansenhovenier1974}. 
The glory feature in Venus's total flux phase curve was later used to rule out potential solutions for the unknown UV absorber \citep{petrova2018glory}.

@@

%--------------------------------------------------------------------------
% Figure 4
%--------------------------------------------------------------------------
\begin{figure*}
    \centering
    \includegraphics{figures/figure-4-scatterplot.pdf}
    \script{create-scatterplot.py}
    \caption{
        Scatter plot for the target star sample, showing stellar effective 
        temperatures and distances.
        %
        The circle sizes represent the angular separation between 
        the star and its planet (in mas) as presented in the target list.
        %
        The colours show $\Delta \varphi$ assuming circular, edge-on orbits
        at a semi-major axis $a$ corresponding to an Earth-like 
        instellation and an IWA of \qty{62}{\mas}.
        %
        Additionally, the colour bar indicates the optical phenomena that 
        can in principle be detected: the phenomena from the bottom of the bar up to the colour of a given circle would be detectable for that planet.
        %
        For example, dark blue circles indicate planets on which most phenomena would be observable because planets that can be observed at the rainbow angle can also be observed at angles exhibiting Rayleigh scattering.
    }
    \label{fig:scatterplot}
\end{figure*}
%--------------------------------------------------------------------------

@@

%--------------------------------------------------------------------------
% Figure 5
%--------------------------------------------------------------------------
%\begin{figure}
%   \centering
%   \includegraphics[width=0.99\columnwidth]{figures/figure-5-orb-grid.pdf}
%   \script{create-figure-5-orb-grid.py}
%   \caption{
%        The phase angles accessible to a direct imaging mission 
%        can be limited by either the orbital inclination angle $i$ or the coronagraphic \IWA{}.
        %
%        If the orbit is close to face-on (top panels), the phase angles probed along the orbit are limited by the inclination angle $i$; for circular orbits, the accessible scattering phases are $\phi \in \qty{90}{\degree} \pm i$. 
        %
%        For orbits close to edge-on (bottom panels) the planet attains all 
%        phases, but the phase angles near \qty{0}{\degree} and \qty{180}{\degree} are not observable due to obscuration by the coronagraph; for circular orbits, the accessible scattering phases are $\phi \in \qty{90}{\degree} \pm \arcsin({\rm IWA}\cdot d_*/a)$.
%    }
%    \label{fig:orb-grid}
%\end{figure}
%--------------------------------------------------------------------------

@@

%--------------------------------------------------------------------------
\begin{figure*}%[t]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure-6-n-over-phase.pdf}
    \script{create-figure-6-n-over-phase.py}
    \caption{
        The cumulative distributions of the most extreme phase angles accessible for different \IWA{} and for randomly inclined, circular orbits
        (solid lines) and randomly inclined, elliptical orbits (dashed lines).
        %
        The top $x$-axis indicates the minimum and the 
        bottom $x$-axis the maximum accessible phase angle.
        %
        These angles are symmetric about quadrature (90 degrees).
        %
        The $y$-axis on the left indicates the number of planetary systems 
        divided by the number of Monte Carlo samples and is thus normalised to 
        the number of systems on the target list.
        %
        The $y$-axis on the right indicates the number of systems, assuming \qty{24}{\percent} of them have an Earth-like planet in their HZ.
    }
    \label{fig:betaallofit}
\end{figure*}
%--------------------------------------------------------------------------

@@

%--------------------------------------------------------------------------
\begin{figure*}%[t]
   \centering
   \includegraphics[width=0.99\textwidth]{figures/figure-7-contrast-over-separation_i=90.pdf}
    \script{create-figure-7-contrast-over-separation.py}
   \caption{
    The contrast (Eq.~\ref{eq:contrast}) and orbital separation 
    of an Earth-like planet with an ocean surface and patchy clouds 
    along its orbit with an assumed inclination angle $i$ of \qty{90}{\degree} around $\alpha$~Cen~A, $\epsilon$~Eri, and Lalande~21185.
    %
    The solid, black line indicates the contrast in unpolarized light, 
    with the contrast at quadrature marked by the black dot.
    %
    The polarized component is indicated by the coloured dots, with the colour representing the phase angle from quadrature.
    %
    The light grey dots show the contrasts at quadrature of the other targets in the star list, and the dashed lines indicate $1$, $2$ and $3$ times the \IWA{} for \hwo.
    }
    \label{fig:contrasts}
\end{figure*}
%--------------------------------------------------------------------------

@@

%--------------------------------------------------------------------------
% Figure 8:
%--------------------------------------------------------------------------
\begin{figure*}
    \centering
    \includegraphics{figures/figure-9-yarnball.pdf}  
    \script{create-figure-8-yarnball.py}
    \caption{
        Examples of the eccentric orbits generated for the stellar sample.
        %
        The orbits are scaled by the Earth-equivalent incident flux distance. 
        %
        The concentric, dashed circles indicate \IWA{}s of \num{20.6}, \num{41.3}, \num{61.9}, and \qty{82.5}{\mas}, corresponding to 1, 2, 3, and 4\,$\lambda / D$, respectively (assuming $\lambda = \qty{600}{\nano\meter}$ and $D = \qty{6}{\meter}$).
        %
        Each system has been scaled such that its HZ has the same radius, 
        hence the variation in the size of the \IWA{}s compared to the orbits.
        It is clear that the \IWA{} can significantly affect the range of 
        observable scattering phases for each orbit.
    }
    \label{fig:ball-o-yarn}
\end{figure*}
%--------------------------------------------------------------------------

%--------------------------------------------------------------------------
% Figure 9:
%--------------------------------------------------------------------------
\begin{figure*}
    \centering
    \includegraphics{figures/figure-9-n-over-iwa.pdf}
    \script{create-figure-9-n-over-iwa.py}
    \caption{
        The number of systems for which key optical phenomena (Rayleigh scattering, rainbow, ocean glint, glories) would be observable as functions of the \IWA.
        %
        Each subplot contains three lines showing the number of systems where the start of the phenomena is observable (dashed lines), where the peak is observable (solid lines), and where the end is observable (dotted lines).
        %
        The $x$-axis at the bottom shows the \IWA{} of the coronagraph in \unit{mas}, and the $x$-axis at the top shows the same \IWA{} converted to $\lambda/D$ for $\lambda = \qty{600}{\nano\meter}$ and $D = \qty{6}{\meter}$.
        %
        The $y$-axis on the right indicates the number of systems where 
        all phenomena would be observable, assuming \qty{24}{\percent} of them have an Earth-like exoplanet in their HZ.
        %
        It is worth noting that the Glory would be observable in only one system: Alpha~Cen~A.
    }
    \label{fig:accessible_phase_angles}
\end{figure*}
%--------------------------------------------------------------------------

>>>>

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is the specular (Fresnel) reflection of light off smooth liquid surfaces. While glint is a familiar phenomenon on Earth, there is only
one other body in the Solar System where it has been observed: 
the Cassini orbiter captured the glint of sunlight reflecting off the
liquid hydrocarbon lakes near Titan's north pole as it was observing
the moon from the nightside \citep{2010GeoRL..37.7104S}.
%
In the total flux, the glint signal is indeed expected to be most prominent at large planetary phase angles.
%
The maximum degree of polarization of the glint occurs at the Brewster angle, which is determined by the refractive indices at the interface; between air and liquid water this occurs at an angle of reflection of \qty{53}{\degree} \citep[i.e., at a planetary phase angle of \qty{106}{\degree}; see, e.g.,][]{Zugger_2010,treesandstam2019}.
%
This angle therefore varies slightly with the wavelength and the composition of the ocean and the atmosphere.
%

In general, the glint on a planet will be strongest at red wavelengths where the Rayleigh scattering in the planet's atmosphere and attenuation through the atmosphere are smallest \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also small \citep{2012ApJ...752L...3C}.
%
The stronger glint signature at longer wavelengths has indeed been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
%
The width across phase angles of the glint feature in total and polarized flux generally
broadens with increasing wave height \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}, which depends strongly on the wind speed over the surface, as described by \citet{CoxMunk1954}.
%

The glint provides evidence of specular reflecting surface bodies, which, taken in the context of the temperature and types of condensates, may be interpreted as evidence for the presence of liquid water.
%
At large phase angles, the forward scattering of light by cloud particles could be confused with glint \citep{Robinson_2010}, which further motivated probing the rainbow angle for context.
%
This confusion can be solved by measuring the intersection of phase curves at various wavelengths in polarized light (the polarized colour changes from blue, through white, to red), which only occurs when the glint contributes to the signal and at a planetary scattering angle depending on the cloud coverage fraction \citep{treesandstam2019}, as shown in Fig.~\ref{fig:bottplot}.
%The cloud coverage fraction on ocean worlds could be constrained with observations spanning the ultraviolet to the near-infrared by finding the 
%angle at which the planet's polarization phase curves at various wavelengths intersect \citep{treesandstam2019}.
%
While the glint feature can only be probed when a planet is at crescent, i.e.\ large phase angles, \citet{trees2022} showed that reflection by a liquid surface body can also be detected with spectropolarimetry, as the glint changes the difference in polarization between continuum wavelengths and absorption bands. This could be applied when the planet is near quadrature, and thus more easily probed when using coronagraphs.
%how use to map, 
%seen in Earthshine
%strong signal for exoplanets
% disentangled from cloud coupled with rainbow
%disentangled from lattitude albedo coupled in polarization

@@

Rainbows, cloudbows, glories and other optical phenomena from light that is 
scattered by atmospheric condensates can dramatically influence the shape of a planet's total flux and polarized phase curves and provide a wealth of information 
about exoplanets \citep{karalidi2012rainbow, stam2008,Bailey2007,2014A&A...566L...1G}.%
\footnote{
    Rainbows occur when light is scattered by rain droplets that are large compared to the wavelength of the light. Cloudbows refer to features occurring at the same phase angles from smaller cloud particles with the spread in colors invisible to the human eye making them appear white \citep[see][for further discussion]{Bailey2007}.  
}
%
In our own Solar System, the polarized back scatter from Titan provided evidence that the moon's reflection was due to an atmospheric haze rather than a surface \citep{zellner1973polarization}, the polarized phase curve of Venus at three wavelengths showed not only that the clouds were $\sim\qty{75}{\percent}$ H\textsubscript{2}SO\textsubscript{4}--water mixture but also constrained the droplet size distribution \citep{hansenhovenier1974}. 
The glory feature in Venus's total flux phase curve was later used to rule out potential solutions for the unknown UV absorber \citep{petrova2018glory}.

@@

% Figure 4
\begin{figure*}
    \centering
    \includegraphics{figures/figure-4-scatterplot.pdf}
    \script{create-figure-4-scatterplot.py}
    \caption{
        Scatter plot for the target star sample, showing stellar effective 
        temperatures and distances.
        %
        The circle sizes represent the angular separation between 
        the star and its planet (in mas) as presented in the target list.
        %
        The colours show $\Delta \varphi$ assuming circular, edge-on orbits
        at a semi-major axis $a$ corresponding to an Earth-like 
        instellation and an IWA of \qty{62}{\mas}.
        %
        Additionally, the colour bar indicates the optical phenomena that 
        can in principle be detected: the phenomena from the bottom of the bar up to the colour of a given circle would be detectable for that planet.
        %
        For example, dark blue circles indicate planets on which most phenomena would be observable because planets that can be observed at the rainbow angle can also be observed at angles exhibiting Rayleigh scattering.
    }
    \label{fig:scatterplot}
\end{figure*}
%--------------------------------------------------------------------------

@@

% Old Figure 5
%\begin{figure}
%   \centering
%   \includegraphics[width=0.99\columnwidth]{figures/figure-5-orb-grid.pdf}
%   \script{create-figure-5-orb-grid.py}
%   \caption{
%        The phase angles accessible to a direct imaging mission 
%        can be limited by either the orbital inclination angle $i$ or the coronagraphic \IWA{}.
        %
%        If the orbit is close to face-on (top panels), the phase angles probed along the orbit are limited by the inclination angle $i$; for circular orbits, the accessible scattering phases are $\phi \in \qty{90}{\degree} \pm i$. 
        %
%        For orbits close to edge-on (bottom panels) the planet attains all 
%        phases, but the phase angles near \qty{0}{\degree} and \qty{180}{\degree} are not observable due to obscuration by the coronagraph; for circular orbits, the accessible scattering phases are $\phi \in \qty{90}{\degree} \pm \arcsin({\rm IWA}\cdot d_*/a)$.
%    }
%    \label{fig:orb-grid}
%\end{figure}

@@

% Figure 5
\begin{figure*}%[t]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure-5-n-over-phase.pdf}
    \script{create-figure-5-n-over-phase.py}
    \caption{
        The cumulative distributions of the most extreme phase angles accessible for different \IWA{} and for randomly inclined, circular orbits
        (solid lines) and randomly inclined, elliptical orbits (dashed lines).
        %
        The top $x$-axis indicates the minimum and the 
        bottom $x$-axis the maximum accessible phase angle.
        %
        These angles are symmetric about quadrature (90 degrees).
        %
        The $y$-axis on the left indicates the number of planetary systems 
        divided by the number of Monte Carlo samples and is thus normalised to 
        the number of systems on the target list.
        %
        The $y$-axis on the right indicates the number of systems, assuming \qty{24}{\percent} of them have an Earth-like planet in their HZ.
    }
    \label{fig:betaallofit}
\end{figure*}
%--------------------------------------------------------------------------

@@

% Figure 6
\begin{figure*}
   \centering
    \includegraphics[width=0.99\textwidth]{figures/figure-6-contrast-over-separation_i=90.pdf}
    \script{create-figure-6-contrast-over-separation.py}
    \caption{
        The contrast (Eq.~\ref{eq:contrast}) and orbital separation 
        of an Earth-like planet with an ocean surface and patchy clouds 
        along its orbit with an assumed inclination angle $i$ of \qty{90}{\degree} around $\alpha$~Cen~A, $\epsilon$~Eri, and Lalande~21185.
        %
        The solid, black line indicates the contrast in unpolarized light, 
        with the contrast at quadrature marked by the black dot.
        %
        The polarized component is indicated by the coloured dots, with the colour representing the phase angle from quadrature.
        %
        The light grey dots show the contrasts at quadrature of the other targets in the star list, and the dashed lines indicate $1$, $2$ and $3$ times the \IWA{} for \hwo.
    }
    \label{fig:contrasts}
\end{figure*}
%--------------------------------------------------------------------------

@@

% Figure 7: Yarn ball
\begin{figure*}
    \centering
    \includegraphics{figures/figure-7-yarnball.pdf}  
    \script{create-figure-7-yarnball.py}
    \caption{
        Examples of the eccentric orbits generated for the stellar sample.
        %
        The orbits are scaled by the Earth-equivalent incident flux distance. 
        %
        The concentric, dashed circles indicate \IWA{}s of \num{20.6}, \num{41.3}, \num{61.9}, and \qty{82.5}{\mas}, corresponding to 1, 2, 3, and 4\,$\lambda / D$, respectively (assuming $\lambda = \qty{600}{\nano\meter}$ and $D = \qty{6}{\meter}$).
        %
        Each system has been scaled such that its HZ has the same radius, 
        hence the variation in the size of the \IWA{}s compared to the orbits.
        It is clear that the \IWA{} can significantly affect the range of 
        observable scattering phases for each orbit.
    }
    \label{fig:ball-o-yarn}
\end{figure*}

% Figure 8
\begin{figure*}
    \centering
    \includegraphics{figures/figure-8-n-over-iwa.pdf}
    \script{create-figure-8-n-over-iwa.py}
    \caption{
        The number of systems for which key optical phenomena (Rayleigh scattering, rainbow, ocean glint, glories) would be observable as functions of the \IWA.
        %
        Each subplot contains three lines showing the number of systems where the start of the phenomena is observable (dashed lines), where the peak is observable (solid lines), and where the end is observable (dotted lines).
        %
        The $x$-axis at the bottom shows the \IWA{} of the coronagraph in \unit{mas}, and the $x$-axis at the top shows the same \IWA{} converted to $\lambda/D$ for $\lambda = \qty{600}{\nano\meter}$ and $D = \qty{6}{\meter}$.
        %
        The $y$-axis on the right indicates the number of systems where 
        all phenomena would be observable, assuming \qty{24}{\percent} of them have an Earth-like exoplanet in their HZ.
        %
        It is worth noting that the Glory would be observable in only one system: Alpha~Cen~A.
    }
    \label{fig:accessible_phase_angles}
\end{figure*}
%--------------------------------------------------------------------------
